Program Optimisations via Hylomorphisms for Extraction of Executable Code

### 2025-07-07 paper/paper.tex
Update
ref MiculanP12

@@

As part of the future improvements, we will study how to mitigate the problem of
setoids. At the moment, we use a short ad-hoc tactic that is able to
automatically discharge many of these proofs in simple settings.  We will study
the more thorough and systematic use of proof automation for proper morphisms.
Generalised rewriting in proofs involving setoids tends to be quite slow, due to
the large size of the terms that need to be rewritten. Sometimes, this size is
hidden in implicit arguments and coercions. We will study alternative
formulations to try to improve the performance of the rewriting tactics
(e.g.\ canonical structures).  Currently, Rocq is unable to inline a number of
trivially inlineable definitions.  We will study alternative definitions, or
extensions to Rocq's code extraction mechanisms to force the full inlining of all
container code that is used in hylomorphisms.  Finally, proving termination
still remains a hurdle. In our framework this reduces to proving that the
anamorphism terminates in all inputs, and we provide a convenient connection to
well-founded recursion. Furthermore, recursive coalgebras compose with natural
transformations, which allows the reuse of a number of core recursive
coalgebras. A possible interesting future line of work is the use of the
approach by Abreu et al.~\cite{AbreuDHJMS23} in combination with ours to improve
code extraction from divide-and-conquer computations whose termination does not
require an external proof.

\newpage

>>>>

\supplement{}

\supplementdetails[subcategory={Source Code}, cite={}, swhid={}]{Software}{https://github.com/dcastrop/coq-hylomorphisms}

@@

As part of the future improvements, we will study how to mitigate the problem of
setoids. At the moment, we use a short ad-hoc tactic that is able to
automatically discharge many of these proofs in simple settings.  We will study
the more thorough and systematic use of proof automation for proper morphisms.
Generalised rewriting in proofs involving setoids tends to be quite slow, due to
the large size of the terms that need to be rewritten. Sometimes, this size is
hidden in implicit arguments and coercions. We will study alternative
formulations to try to improve the performance of the rewriting tactics
(e.g.\ canonical structures).  Currently, Rocq is unable to inline a number of
trivially inlineable definitions.  We will study alternative definitions, or
extensions to Rocq's code extraction mechanisms to force the full inlining of all
container code that is used in hylomorphisms.  Following work by Miculan and
Paviotti~\cite{MiculanP12} an interesting next step would be to embed the notion
of hylomorphism into a type theory with a notion of distributed computations
(e.g. ML5), the resulting extracted code would be then implemented in a
distributed programming language (e.g. Erlang).
Finally, proving termination still remains a hurdle. In our framework this
reduces to proving that the anamorphism terminates in all inputs, and we provide
a convenient connection to well-founded recursion. Furthermore, recursive
coalgebras compose with natural transformations, which allows the reuse of a
number of core recursive coalgebras. A possible interesting future line of work
is the use of the approach by Abreu et al.~\cite{AbreuDHJMS23} in combination
with ours to improve code extraction from divide-and-conquer computations whose
termination does not require an external proof.
